---
title: "Introduction to RNAseq"
author: Nicole Black
date: 07/24/22
---

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/talhaaccount/Desktop/Qbio_490_R/qbio_490_talha/week6_rna/")) 
```

```{r}
# install, only if needed, and load packages (SummarizedExperiment and TCGAbiolinks)
```

Make sure you have access to rna_se. If it is not already in your workspace (ie already queried, downloaded, and prepared), let Nicole know NOW so we can get you the data a different way.

###### (1) Summarized Experiment Basics

Let's take a look at rna_se (RNA Summarized Experiment). Call head() and str(). Also explore the data frame by double clicking on it from the Environment panel (or calling "View(rna_se)").

```{r Exercise 1.1}
# write code here
head(rna_se)
```

Use the dim() function to find the dimensions of the colData, rowRanges, and unstranded dataframes. Record them below:

  ~ colData (1226 rows, 87 cols):
  ~ rowRanges: 60660 10
  ~ unstranded: 60660 1226

```{r Exercise 1.2}
# write code here
dim(rna_se@colData)
dim(rna_se@rowRanges@elementMetadata)   #DID THIS WRONG
dim(rna_se@assays@data$unstranded)

```

Run the following code and fill out a description of the rows/columns of our data frames.You can use this as a cheat sheet moving forward.
  ~ colData
        - Rows: Patients
        - Columns: info about those patients 
  ~ rowRanges
        - Rows: Genes
        - Columns: info about those genes
  ~ unstranded (in assays)
        - Rows: genes
        - Columns: patients 
  
```{r Exercise 1.3}
# colData: 
rownames()
colnames()

# rowRanges:
str()
# what are the rows... use your logic

# for unstranded, don't call anything, just use your logic from exercise 1.2 and and think it through!
```

###### (2) ColData
# rna_se@colData is essentially our clinical dataframe but for rna. it gives further informaton about each patient

View the first 15 values in the age_at_index column.

```{r Exercise 2.1}
# write code here
(rna_se@colData$age_at_index[1:15])
```

Call summary on the age_at_index column. Record the output below:
    Min.: 26.00
    1st Qu.: 48.00
    Median: 58.00
    Mean: 58.27
    3rd Qu.: 67.00
    Max.: 90.00
    NA's: 1
    
```{r Exercise 2.2}
# write code here
summary(rna_se@colData$age_at_index)
```

Create a data frame called rna_clinical that contains data from colData, except for the NA patient and their data.

```{r Exercise 2.3}
#creating a mask, that will then make a new df called rna_clinical without NA's
mask <- ifelse(is.na(rna_se@colData$age_at_index),F,T)
rna_clinical <- rna_se@colData[mask, ]
```

Use the as.data.frame() function to change this from an S4 object to a data frame object.

```{r Exercise 2.4}
rna_clinical <- as.data.frame(rna_clinical)
```

Make a mask and subset out the treatments column.

```{r Exercise 2.5}
treatments_mask <- 
rna_clinical <- 
```

Save rna_clinical to your local machine

```{r Exercise 2.6}
write.csv(rna_clinical, "/PATH/TO/FILE/brca_rna_clincial_data.csv", row.names = TRUE)
```

###### (3) RowRanges

Call head() on rowRanges@elementMetadata:

```{r Exercise 3.1}
# write code here
```

Create a data frame called rna_genes that contains data from elementMetadata.

```{r Exercise 3.2}
rna_genes <- 
```

Use the as.data.frame() function to change this from an S4 object to a data frame object

```{r Exercise 3.3}
rna_genes <- as.data.frame(FILL IN HERE)
```

Save rna_genes to your local machine.

```{r Exercise 3.4}
write.csv(rna_genes, "/PATH/TO/FILE/brca_rna_gene_data.csv", row.names = FALSE)
```

###### (4) Assays

What are the names of the 6 assays?

```{r Exercise 4.1}
# access the assays here
```

Create a new data frame called "rna_counts" that contains the data from unstranded. Recall the patient with NA age data? We need to exclude that patient. We can use the same mask to do this.

```{r Exercise 4.2}
rna_counts <- FILL IN HERE
```

Fetch the data for the first 5 patients and the first 15 rows.

```{r Exercise 4.3}
# write code here
```

Save rna_counts to your local machine (warning: this will likely take a while!)

```{r Exercise 4.4}
write.csv(rna_counts, "/PATH/TO/FILE/brca_rna_count_data.csv", row.names = TRUE)
```
